Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2018-05-01 manuscript/manuscript.tex
edited figure captions

\begin{figure}[tb]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_A5e-15_dataset11.pdf}
	\caption{Optimal statistic for a simulated PTA dataset containing an injected GWB. 
			We compare the values found using a fixed-noise analysis 
			with noise parameters taken from 
			individual and common noise analyses 
			(dashed blue lines and solid orange lines, respectively) 
			to the values found by marginalizing over 
			the individual pulsars' red noise parameters (green histograms). 
			The dotted vertical line indicates the injected value, $\Agw^2 = 2.5 \times 10^{-29}$. 
			The fixed-noise analysis 
			using the individual noise results underestimates $\Agw$, 
			while the fixed-noise analysis using the common noise results 
			and the noise-marginalized analysis more accurately recover $\Agw$.}
	\label{fig:os_dataset_sample}
\end{figure}

In Fig.~\ref{fig:os_datasetstats} we show the optimal statistic 
for 200 different realizations of a GWB with $\Agw = 5\times10^{-15}$ 
computed using the three techniques described above. 
Using the noise values from individual noise analyses 
systematically underestimates the strength of the GWB, 
while using the noise values from a common noise analysis 
more accurately recovers the injected value. 
The fixed-noise analysis using the individual noise results finds 
$\hat{A}^2 = (7.9 \pm 6.8) \times10^{-30}$ and $\rho = 2.3 \pm 1.5$, 
averaging over realizations of the GWB. 
The fixed-noise and noise-marginalized analyses 
using the common noise results both give 
$\hat{A}^2 = (2.4\pm1.2)\times10^{-29}$ and $\rho = 4.1 \pm 1.7$.

\begin{figure}[tb]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_A5e-15.pdf}
	\caption{Optimal statistic and SNR for 200 simulated datasets 
			containing an injected GWB. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			We show the values found by fixing the red noise 
			based on the results from individual and common noise analyses 
			(blue and orange lines, respectively) 
			as well as the mean optimal statistic and mean SNR found by marginalizing over 
			the red noise parameters using the results of a common noise analysis (green lines). 
			Using the red noise values from individual noise analyses leads to values of $\hat{A}^2$ 
			that are significantly lower than the injected value, 
			while using the values from a common noise analysis gives values of $\hat{A}^2$ 
			that are closer to the injected value.}
	\label{fig:os_datasetstats}
\end{figure}

@@

The optimal statistic is particularly well-suited to compare multiple spatial correlation relations 
because using a different spatial correlation only requires changing the ORF 
in Eq.~\eqref{eq:Sab}. 
\citet{thk+2016} demonstrated how the optimal statistic can be altered to fit for 
multiple spatial correlations at once in order to mitigate common noise sources such as 
clock error and ephemeris error. 
Here we compute the optimal statistic with monopole and dipole spatial correlations 
for the same simulated data sets as in the previous section in order to determine 
how well we can distinguish a GW background from a monopole or dipole signal. 
For a monopole signal, the ORF becomes simply
$\Gamma(\theta_{ab}) = 1$, 
while for a dipole signal, the ORF becomes
$\Gamma(\theta_{ab}) = \cos\theta_{ab}$.

@@

We generated \sv{XXX} scrambled ORFs 
using a Monte Carlo algorithm. 
We required the scrambled ORFs to have $\bar{M} < 0.2$ 
with respect to the true ORF and each other 
in order to minimize the overlap between the ORFs. 
%Figure~\ref{fig:skyscrambles_dataset_sample} compares the noise-marginalized optimal statistic 
%for a particular data set to the distribution of noise-marginalized optimal statistics obtained using sky scrambles. 
%For this data set, the noise-marginalized mean optimal statistic is 
%$1.9\times10^{-29}$ and the mean SNR is 1.1. 
%The distributions of the noise-marginalized optimal statistic and SNR obtained using the scrambled ORFs 
%are both centered around zero, 
%and for 24 out of the 210 sky scrambles, the SNR is above the true value of $1.1$ ($p=0.11$). 
%Figure~\ref{fig:skyscrambles} shows the $p$-values for 210 sky scrambles 
%using 425 realizations of the stochastic background. 
%For 62\% of simulations the $p$-value is $<0.1$, and for 29\% of simulations the $p$-value is $<0.01$.
%\begin{figure}[htb]
%	\includegraphics[width=0.9\columnwidth]{plots/skyscrambles_dataset50.pdf}
%	\caption{Comparison between the noise-marginalized mean optimal statistic and mean SNR 
%			with and without sky scrambles for a simulated dataset 
%			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
%			Using the true ORF, we found an SNR of 1.1, while only 24 of the 210 analyses using scrambled ORFs 
%			found an SNR $\geq 1.1$ ($p=0.11$).}
%	\label{fig:skyscrambles_dataset_sample}
%\end{figure}
%\begin{figure}[htb]
%	\includegraphics[width=0.9\columnwidth]{plots/os_skyscrambles.pdf}
%	\caption{Distribution of $p$-values for 425 different realizations of the stochastic background. 
%			For each realization, the $p$-value is the fraction of sky scrambled ORFs that gives a 
%			noise-marginalized mean SNR greater than the mean SNR using the correct ORF.}
%	\label{fig:skyscrambles}
%\end{figure}

>>>>

\begin{figure}[tb]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_A5e-15_dataset11.pdf}
	\caption{Optimal statistic 
			for a simulated PTA dataset containing a GWB with $\Agw = 5\times10^{-15}$. 
			The fixed-noise analysis using the individual noise values (dashed blue lines) 
			systematically underestimates $\hat{A}^2$, while the 
			fixed-noise analysis using the common noise values (solid orange lines) 
			and the noise-marginalized analysis (green histograms) 
			more accurately recover $\Agw$.}
	\label{fig:os_dataset_sample}
\end{figure}

In Fig.~\ref{fig:os_datasetstats} we show the optimal statistic 
for 200 different realizations of a GWB with $\Agw = 5\times10^{-15}$ 
computed using the three techniques described above. 
For the noise-marginalized analysis, we plot 
the mean values of $\hat{A}^2$ and $\rho$. 
Using the noise values from individual noise analyses 
systematically underestimates the strength of the GWB, 
while using the noise values from a common noise analysis 
more accurately recovers the injected value. 
The fixed-noise analysis using the individual noise results finds 
$\hat{A}^2 = (7.9 \pm 6.8) \times10^{-30}$ and $\rho = 2.3 \pm 1.5$, 
averaging over realizations of the GWB. 
The fixed-noise and noise-marginalized analyses 
using the common noise results both give 
$\hat{A}^2 = (2.4\pm1.2)\times10^{-29}$ and $\rho = 4.1 \pm 1.7$.

\begin{figure}[tb]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_A5e-15.pdf}
	\caption{Optimal statistic and SNR for 200 simulated datasets 
			containing a GWB with $\Agw = 5\times10^{-15}$. 
			The fixed-noise analysis using the individual noise values (blue) 
			systematically underestimates $\hat{A}^2$, while both the 
			fixed-noise analysis using the common noise values (orange) 
			and the noise-marginalized analysis (green) 
			accurately recover $\Agw$.}
	\label{fig:os_datasetstats}
\end{figure}

@@

The optimal statistic is particularly well-suited 
to compare multiple spatial correlation relations 
because using a different spatial correlation only requires changing the ORF 
in Eq.~\eqref{eq:Sab}. 
\citet{thk+2016} demonstrated how the optimal statistic can be altered to fit for 
multiple spatial correlations at once in order to mitigate common noise sources such as 
clock error and ephemeris error. 
Here we take a different approach -- rather than simultaneously fitting 
for signals with different spatial correlations, 
we look at how well we can distinguish between different spatial correlations 
by computing the optimal statistic with monopole and dipole spatial correlations 
for the same simulated data sets as in the previous section. 
For a monopole signal, the ORF becomes simply
$\Gamma_{ab} = 1$, 
while for a dipole signal, the ORF becomes
$\Gamma_{ab} = \cos\theta_{ab}$.

@@

We generated 725 scrambled ORFs 
using a Monte Carlo algorithm. 
We required the scrambled ORFs to have $\bar{M} < 0.2$ 
with respect to the true ORF and each other 
in order to minimize the overlap between the ORFs. 
For each simulation, 
we computed the noise-marginalized mean optimal statistic and mean SNR 
for each scrambled ORF, and compared to the values found using the true ORF. 
Figure~\ref{fig:skyscrambles_dataset_sample} shows the results of a sky scramble analysis 
for a sample dataset with $\Agw=5\times10^{-15}$. For this particular realization, 
the noise-marginalized optimal statistic 
using the true ORF found $\hat{A}^2=( \pm ) \times 10^{-29}$ 
and $\rho = \pm $. 
None of the 725 scrambled ORFs resulted in a mean SNR greater than 4.8 ($p < 0.0014$).
\begin{figure}[htb]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_scrambled_dataset11.pdf}
	\caption{Comparison between the noise-marginalized mean optimal statistic and mean SNR 
			with and without sky scrambles for a simulated dataset 
			containing an injected GWB with $\Agw = 5\times10^{-15}$. 
			The true ORF gave a mean SNR of 4.8. 
			None of the 725 scrambled ORFs gave a mean SNR this large ($p < 0.0014$).}
	\label{fig:skyscrambles_dataset_sample}
\end{figure}

In Fig.~\ref{fig:skyscrambles_stats}, we plot the distribution of $p$-values of the 725 sky scrambles 
for 200 realizations of the GWB.

%Figure~\ref{fig:skyscrambles_dataset_sample} compares the noise-marginalized optimal statistic 
%for a particular data set to the distribution of noise-marginalized optimal statistics obtained using sky scrambles. 
%For this data set, the noise-marginalized mean optimal statistic is 
%$1.9\times10^{-29}$ and the mean SNR is 1.1. 
%The distributions of the noise-marginalized optimal statistic and SNR obtained using the scrambled ORFs 
%are both centered around zero, 
%and for 24 out of the 210 sky scrambles, the SNR is above the true value of $1.1$ ($p=0.11$). 
%Figure~\ref{fig:skyscrambles} shows the $p$-values for 210 sky scrambles 
%using 425 realizations of the stochastic background. 
%For 62\% of simulations the $p$-value is $<0.1$, and for 29\% of simulations the $p$-value is $<0.01$.
%\begin{figure}[htb]
%	\includegraphics[width=0.9\columnwidth]{plots/skyscrambles_dataset50.pdf}
%	\caption{Comparison between the noise-marginalized mean optimal statistic and mean SNR 
%			with and without sky scrambles for a simulated dataset 
%			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
%			Using the true ORF, we found an SNR of 1.1, while only 24 of the 210 analyses using scrambled ORFs 
%			found an SNR $\geq 1.1$ ($p=0.11$).}
%	\label{fig:skyscrambles_dataset_sample}
%\end{figure}
%\begin{figure}[htb]
%	\includegraphics[width=0.9\columnwidth]{plots/os_skyscrambles.pdf}
%	\caption{Distribution of $p$-values for 425 different realizations of the stochastic background. 
%			For each realization, the $p$-value is the fraction of sky scrambled ORFs that gives a 
%			noise-marginalized mean SNR greater than the mean SNR using the correct ORF.}
%	\label{fig:skyscrambles}
%\end{figure}
